# NOAA Storm Data Analysis
#### Carol Tao
#### Sept.3, 2020

# Analysis of NOAA data to assess the health and economic impacts of severe storms in the United States

## 1. Description of Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis. Source: Coursera Assignment Description

## 2. Synopsis
NOAA is the National Oceanic and Atmospheric Administration in the United States. They aim to provide accurate data and cutting edge research in their field. Today we are looking at a public dataset that maintains weather data per storm event dating back over 50 years. These data can show us which weather events have occurred and what implications these events can have on the safety and well being of the surrounding communities. We can see the damage to property, people, and crops, all organized by storm event type.

As you’ll see in these data analyzed below, there were some events that proved to be more dangerous than others. The weather event that causes the most harm to public health is Tornadoes. They have shown in the graphs above to be the largest cause of fatalities and injuries due to weather events in the United States. When strictly talking about economic damages, the events that have caused the most damage are Flood, Drought and Tornado, but for different reasons. For example the biggest risk to crops is a hail event, whereas the biggest threat to properties are tornados.

## 3. Data Processing
3.1
Below are the packages I used in this analysis
```{r packages}
library(dplyr)
library(ggplot2)
library(knitr)
library(cowplot)
```

Download the data from the internet and read the data
```{r read_data}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = "~/Desktop/R_Files/Storm_Analysis/storm.csv", method = "curl")

##check if the file is downloaded
list.files("~/Desktop/R_Files/Storm_Analysis")

##collect the infomation on when the file is downloaded
dateDownloaded <- date()
dateDownloaded

all_data <- read.csv("storm.csv")
```

3.2 Subset the data to only include the columns we need
```{r select}
storm <- select(all_data, 'EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')
head(storm)
```

3.3. Clean up data set and prepare data for analysis
Prepare property damage data and crop data to allow exponents to show correctly in your tidy data set and figures.

Property Exponents
```{r property}
storm$PROPEXP[storm$PROPDMGEXP == "K"] <- 1000
storm$PROPEXP[storm$PROPDMGEXP == "M"] <- 1e+06
storm$PROPEXP[storm$PROPDMGEXP == ""] <- 1
storm$PROPEXP[storm$PROPDMGEXP == "B"] <- 1e+09
storm$PROPEXP[storm$PROPDMGEXP == "m"] <- 1e+06
storm$PROPEXP[storm$PROPDMGEXP == "0"] <- 1
storm$PROPEXP[storm$PROPDMGEXP == "5"] <- 1e+05
storm$PROPEXP[storm$PROPDMGEXP == "6"] <- 1e+06
storm$PROPEXP[storm$PROPDMGEXP == "4"] <- 10000
storm$PROPEXP[storm$PROPDMGEXP == "2"] <- 100
storm$PROPEXP[storm$PROPDMGEXP == "3"] <- 1000
storm$PROPEXP[storm$PROPDMGEXP == "h"] <- 100
storm$PROPEXP[storm$PROPDMGEXP == "7"] <- 1e+07
storm$PROPEXP[storm$PROPDMGEXP == "H"] <- 100
storm$PROPEXP[storm$PROPDMGEXP == "1"] <- 10
storm$PROPEXP[storm$PROPDMGEXP == "8"] <- 1e+08

storm$PROPEXP[storm$PROPDMGEXP == "+"] <- 0
storm$PROPEXP[storm$PROPDMGEXP == "-"] <- 0
storm$PROPEXP[storm$PROPDMGEXP == "?"] <- 0

storm$PROPDMGVAL <- storm$PROPDMG * storm$PROPEXP
```

Crop Exponents
```{r crop}
storm$CROPEXP[storm$CROPDMGEXP == "M"] <- 1e+06
storm$CROPEXP[storm$CROPDMGEXP == "K"] <- 1000
storm$CROPEXP[storm$CROPDMGEXP == "m"] <- 1e+06
storm$CROPEXP[storm$CROPDMGEXP == "B"] <- 1e+09
storm$CROPEXP[storm$CROPDMGEXP == "0"] <- 1
storm$CROPEXP[storm$CROPDMGEXP == "k"] <- 1000
storm$CROPEXP[storm$CROPDMGEXP == "2"] <- 100
storm$CROPEXP[storm$CROPDMGEXP == ""] <- 1
storm$CROPEXP[storm$CROPDMGEXP == "?"] <- 0
storm$CROPDMGVAL <- storm$CROPDMG * storm$CROPEXP
```

## 4. Results
### 4.1. Question 1
Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

#### 4.1.1 Plot total fatalities against weather events
```{r fatalities_weather}
## calculate total fatalities for each weather and choose the top 10 hightest fatalities
storm_evtype_1 <- aggregate(FATALITIES ~EVTYPE, data = storm, FUN = sum ) 
fatalities <- arrange(storm_evtype_1, desc(FATALITIES))
top_10_1 <- fatalities[1:10,]
top_10_1$EVTYPE <- factor(top_10_1$EVTYPE, levels = fatalities$EVTYPE)

## plot
g3 <- ggplot(top_10_1, aes(x = EVTYPE, y = FATALITIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event Type") + ylab("Fatalities") + ggtitle("Fatalities by Weather")

```

#### 4.1.2 Plot total injuries against weather events
```{r injuries_weather}
## calculate total injuries for each weather and choose the top 10 hightest injuries
storm_evtype_2 <- aggregate(INJURIES ~EVTYPE, data = storm, FUN = sum ) 
injuries <- arrange(storm_evtype_2, desc(INJURIES))
top_10_2 <- injuries[1:10,]
top_10_2$EVTYPE <- factor(top_10_2$EVTYPE, levels = injuries$EVTYPE)

g4 <- ggplot(top_10_2, aes(x = EVTYPE, y = INJURIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event Type") + ylab("Injuries") + ggtitle("Injuries by Weather")

## combine the two graphs together using the cowplot package
plot_grid(g3,g4,ncol = 2, nrow = 1)
```

Conclusion: The weather event that causes the most harm to public health is Tornadoes. They have shown in the graphs above to be the largest cause of fatalities and injuries due to weather events in the United States.

### 4.2. Question 2
Across the United States, which types of events have the greatest economic consequences?

#### 4.2.1 Plot total property damages against weather events
```{r property_weather}
## calculate total property damages for each weather and choose the top 10 hightest damages
storm_evtype_3 <- aggregate(PROPDMG ~EVTYPE, data = storm, FUN = sum )
property<- arrange(storm_evtype_3, desc(PROPDMG))
top_10_3 <- property[1:10,]
top_10_3$EVTYPE <- factor(top_10_3$EVTYPE, levels = property$EVTYPE)

g1 <- ggplot(top_10_3, aes(x = EVTYPE, y = PROPDMG)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event Type") + ylab("Property Damage") + ggtitle("Property damage by Weather")

```

#### 4.2.2 Plot total crop damages against weather events
```{r crop_weather}
## calculate total property damages for each weather and choose the top 10 hightest damages
storm_evtype_4 <- aggregate(CROPDMG ~EVTYPE, data = storm, FUN = sum )
crop<- arrange(storm_evtype_4, desc(CROPDMG))
top_10_4 <- crop[1:10,]
top_10_4$EVTYPE <- factor(top_10_4$EVTYPE, levels = crop$EVTYPE)

g2 <- ggplot(top_10_4, aes(x = EVTYPE, y = CROPDMG )) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event Type") + ylab("Crop Damage") + ggtitle("Crop damage by Weather")

## combine the two graphs together using the cowplot package
plot_grid(g1,g2,ncol = 2, nrow = 1)
```

Conclusion:  When strictly talking about economic damages, the events that have caused the most damage are Flood, Drought and Tornado, but for different reasons. For example the biggest risk to crops is a hail event, whereas the biggest threat to properties are tornados.


